Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-09-08 Chapter_1/04_Manuscript/Manuscript_2021-07-14.tex
Various updates

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
% Importance of Dispersal & Connectivity
Dispersal of individuals is a crucial pre-requisite for long-term species
persistence. However, the ability to disperse is contingent a sufficient degree
of landscape connectivity, which is why the identification and preservation of
movement corridors that promote connectivity has become a task of extraordinary
importance for conservation authorities.

% Quantifying connectivity
During the past two decades, ecologists have largely relied on graph-based
methods, such as least-cost analysis or circuit theory, to investigate
connectivity. Despite their popularity, both methods make several restrictive
assumptions that limit their usefulness in reality. Least-cost analysis, for
instance, assumes that animals move towards a pre-conceived endpoint and choose
a cost-minimizing route accordingly. Circuit theory, in contrast, presumes that
animals move according to a random walk without any directional persistence. To
overcome these shortcomings, dispersal simulations from individual-based
movement models have been proposed, yet a unified framework to simulate
dispersal and assess connectivity is lacking.

% What we do
Here, we examine the use of integrated step selection analysis as a simple but
powerful framework for fitting individual-based movement models and simulating
dispersal. For this, we analyse data of 16 dispersing wild dogs (\textit{Lycaon
pictus}) and parametrize a fully mechanistic movement model describing how
dispersers move through the available landscape. We then employ the model to
simulate 80'000 dispersing wild dogs moving across the world's largest
transboundary conservation area. Finally, we use the simulated trajectories to
derive three complementary connectivity maps, each focusing on a different
aspect of connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors,
and a map of inter-patch connectivity, depicting the presence or absence of
functional links between distinct habitat patches.

% Relevance
We conclude that simulations from integrated step selection functions indeed
offer a powerful alternative to traditional connectivity modeling techniques and
requires fewer unrealistic assumptions about dispersal. Nevertheless, some
non-trivial modeling decisions such as the number and duration of simulated
dispersal events need to be considered.

@@

\section{Introduction}
% Importance of Dispersal & Connectivity
\subsection{Dispersal \& Connectivity (90\%)}
Successful dispersal of individuals requires a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become a fundamental task in
conservation science \citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Pinpointing
relevant dispersal hotspots not only necessitates information on movement
behavior and knowledge about factors that limit dispersal \citep{Baguette.2013,
Vasudev.2015, Hofmann.2021}, but also an adequate connectivity model
\citep{Diniz.2020}. To date, two of the most prominent connectivity models used
to infer movement corridors are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based approaches that estimate conductance of the
landscape based on estimated habitat permeability for the focal species
\citep{Zeller.2012}.

% Issues with Least-Cost Paths
\subsection{Issues with LCPs and CT (90\%)}
Despite their intuitive nature and ease of use, least-cost analysis and circuit
theory make several rigorous assumptions that are hardly ever met in reality
\citep{Diniz.2020}. With LCP analysis, for instance, a least costly route always
exists, even if associated movement costs are unreasonably high and will never
be incurred by a dispersing individual. The method also presumes that animals
move towards a pre-conceived endpoint and choose a cost-minimizing route
accordingly, which further requires perfect knowledge about landscape and
associated movement costs. While this may be a reasonable assumption for
migrating animals, dispersers usually move into unfamiliar territory and are
therefore unaware of associated movement costs and potential endpoints
\citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Circuit theory relaxes these
assumptions, yet requires that animals move according to a random walk,
entailing that directional biases cannot be rendered \citep{Diniz.2020}.
Nevertheless, directional movements are regularly observed \citep{Bovet.1991,
Schultz.2001}, particularly in dispersing individuals \citep{Cozzi.2020,
Hofmann.2021}. Some of these deficiencies can be addressed using intermediate
methods, such as (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016}, yet a certain
degree of arbitrariness remains. In addition, neither of these methods is
capable of rendering the temporal dimension of dispersal \citep{Diniz.2020},
such that statements about the expected duration required to traverse a certain
corridor are impossible.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To overcome the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
individual dispersal trajectories are simulated based according to movement
rules that determine how individuals move across or interact with the prevailing
landscape \citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004,
Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020, Vasudev.2021}. To quantify landscape connectivity
from simulated trajectories, as set of connectivity metrics, such as inter-patch
connectivity and traversal frequency, to reveal major dispersal corridors
\citep{Kanagaraj.2013, BastilleRousseau.2018, Hauenstein.2019, Zeller.2020} can
be calculated. However, while IBMMS can be employed to overcome any of the
shortcomings intrinsic to LCPs and CT, they are subject to a vast amount of
subjective, non-trivial modeling decisions. Moreover, they can be challenging to
fit and require vast amounts of movement data, ideally collected during
dispersal \citep{Diniz.2020}. Consequently, alternative methods that require
fewer modeling decisions and are straight forward to apply are desirable.

@@

% Previous Paper
\subsection{Previous Paper (90\%)}
We previously assessed landscape connectivity for dispersing African wild dogs
within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a habitat selection model and predicted a permeability
surface that we used to compute least-cost paths and corridors between protected
areas. We now expand on this knowledge and use ISSFs to develop a more detailed,
mechanistic movement model of dispersing wild dogs (\Cref{GraphicalAbstract}).
We employ the model to simulate dispersers moving across the KAZA-TFCA and
generate three distinct connectivity maps, each shedding light onto a different
aspect of connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors and
bottlenecks, and a map of inter-patch connectivity, depicting the presence or
absence of functional links between national parks as well as the average
dispersal duration required to realize those links. With this work, we exemplify
how ISSFs can be utilized for dispersal simulations and we discuss several
benefits of this approach over traditional connectivity modeling techniques such
as least-cost analysis and circuit theory. Most importantly, simulations based
on ISSFs provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addition, by
generating proper dispersal trajectories, network theory can be applied to
calculate network metrics that are pertinent to connectivity analysis.

@@

\subsection{GPS Relocation Data (90\%)}
GPS data of dispersing wild dogs was obtained as described in \citep{Cozzi.2020,
Hofmann.2021}. We collected data on 7 female and 9 male dispersing coalitions
between 2011 and 2019 from a free-ranging population inhabiting the Moremi
National Park in northern Botswana. In some instances, exact dispersal dates
were known from field observations, in other cases we determined dispersal
phases using the net-squared displacement metric. Net squared displacement
measures the squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. Thus, dispersal was deemed to have started when
an individual left its natal home range and ended once individuals became
sedentary again. Because behavior during dispersal is more pertinent to
landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data
that was collected during residency. During dispersal, collars were programmed
to record a GPS fix every four hours and regularly transmitted over the Iridium
satellite system. As previous research found no differences in behaviors of
females and males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did
not distinguish between sexes. After collection, we converted GPS coordinates (n
= 4'169) to steps, where each step represented the straight-line distance
traveled by an individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule (\( \pm \) 15
minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting water layers thus temporally aligned with
each dispersal event, ensuring minimal temporal lag between observed movements
and corresponding water layer. We furthermore computed a proxy for human
influence, rendering anthropogenic pressures stemming from human-density,
agricultural sites, and roads. All spatial layers were coarsened or interpolated
to a target resolution of 250 m by 250 m. A detailed description of the
preparation of each habitat covariate is given in \cite{Hofmann.2021}.

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps, such that a realized step plus its 24 random steps formed
a stratum that received a unique identifier. As suggested by \cite{Avgar.2016},
we generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale = 6'308, shape = 0.37). Along each step, we
extracted and averaged spatial covariates using the {\tt velox} package
\citep{Hunziker.2021}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each observed and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| > 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

@@

\subsection{Dispersal Simulation (80\%)}
We used the most parsimonious movement model to simulate 80'000 virtual
dispersers moving across the KAZA-TFCA. The simulation resembled an inverted
ISSF and was set up as follows. (1) We defined a random source point and assumed
a random initial orientation of the animal. (2) Departing from the source point,
we generated 25 random steps by sampling turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from our fitted gamma
distribution (similar to the input data, each random step represented the
straight line movement within 4 hours). To prevent unreasonably large steps, we
capped sampled step lengths to a maximum of 35 km, which corresponded to the
farthest distance ever traveled within 4 hours by one of our dispersers. (3)
Along each random step, we extracted and averaged habitat covariates and we
calculated movement covariates. To ensure compatible scales, we standardized
extracted values using the same parameters applied to our input data. (4) We
applied the parametrized movement model to predict the selection score \(w(x)\)
for each step and we translated predicted scores into probabilities using
\Cref{EQ2}. (5) We sampled one of the random steps based on predicted
probabilities and determined the animal's new position. We repeated steps (2) to
(5) until 2'000 steps were realized, implying a total 160 Mio. simulated steps.

@@

\subsection{Source Points (90\%)}
We released 80'000 virtual dispersers from 80'000 unique source points
distributed across the study area. 50'000 virtual dispersers were released from
randomly selected source points within contiguous protected areas that were
large enough to sustain viable wild dog populations (\(>\) 700
km\textsuperscript{2}; \citealp{Pomilia.2015}) and another 30'000 were released
at random locations inside the 100 km wide buffer to render potential immigrants
into the study system. By distributing source points randomly, the number of
source points per km\textsuperscript{2} was approximately equal within protected
areas and the buffer zone.

% \begin{figure}[htbp]
%   \begin{center}
%     \includegraphics[width = \textwidth]{99_SourceAreas.png}
%     \caption{(a) Different source areas from which we released virtual
%     dispersers. We only considered contiguous protected areas (national parks,
%     game reserves, and forest reserves) that were larger than 700
%     km\textsuperscript{2} (green). This area corresponds to the average home
%     range requirement for viable wild dog populations \citep{Pomilia.2015}. To
%     render potential immigrants into the study system, we also initiated
%     dispersers within a buffer zone (blue) surrounding the main study area. (b)
%     Source points from which dipsersers were released. 50'000 dispersers were
%     released from the main study area (green dots) and another 30'000 dispersers
%     within the virtual buffer (blue dots).}
%     \label{SourcePoints}
%   \end{center}
% \end{figure}

@@

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers \citep{Hauenstein.2019, Peer.2008}. For this,
we rasterized all simulated trajectories and tallied them into a single map. If
the same trajectory crossed a raster-cell twice, we only counted it once,
thereby mitigating potential biases caused by individuals that were trapped and
moved in circles. To achieve high performance rasterization, we used the
R-package {\tt terra} \citep{Hijmans.2020}.

@@

\subsection{Inter-Patch Connectivity (80\%)}
We assessed inter-patch connectivity between national parks located in our study
area to examine functional links between distinct patches in the KAZA-TFCA. The
decision to focus on national parks was purely out of simplicity and does not
imply that connections between other regions are impossible. In fact, the same
process could easily be expanded to include other protected areas. To quantify
inter-patch connectivity, we computed the relative frequency at which dispersers
originating from one national park successfully moved into another national
park. Successful movement was said to be achieved if the individuals' trajectory
intersected with the corresponding national park at least once. We also recorded
the number of steps required until the first intersection with the respective
national park. This allowed us to determine \textit{if} and \textit{how often}
dispersers moved between certain national parks, as well as \textit{how long}
dispersers had to move to realize those connections.

@@

Assuming that all other covariates are held at their means, the habitat kernel
reveals that dispersing wild dogs avoid water but prefer its proximity.
Similarly, dispersers avoid areas that are covered by woodlands, yet prefer
regions covered by shrublands or grasslands. Finally, dispersers avoid movement
through landscapes that are dominated by humans. Effect sizes are strong and,
except for effect of \textsf{distance to water}, statistically clear on the 5\%
significance level.

@@

\subsection{Dispersal Simulation (80\%)}
On a machine with an octacore AMD Ryzen 7 2700X processor (8 x 3.6 GHz) and 64
GB of RAM, a batch of 1'000 simulated dispersers moving over 2'000 steps
required 90 minutes to compute (\(\mu = 88.90\), \(\sigma = 1.87\)).
Consequently, the simulation of all 80'000 dispersers (160 Mio. steps)
terminated after 120 hours (i.e. five days). Comparable simulations will be
substantially faster for smaller study areas and lower resolution covariates, as
the covariate extraction from large and high-resolution rasters was
computationally the most demanding task. Out of the 50'000 dispersers initiated
inside the main source area \Cref{SourcePoints}(a), only 4.5\% eventually hit a
map boundary, suggesting that biases due to boundary effects are moderate. In
contrast, 78\% of the 30'000 dispersers originating from the buffer zone
eventually hit a map boundary, yet this was to be expected since many of those
dispersers originated from source points located close to the map boundary.

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
% Dispersal & Connectivity
Dispersal is an important process that allows species to avoid inbreeding,
colonize new habitats and reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and preservation of dispersal corridors has become a
task of extraordinary importance for conservation authorities.

% Quantifying Connectivity
Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their usefulness in reality. Individual-based dispersal simulations have
been proposed to address these shortcomings, yet due to the sheer amount of
non-trivial modeling decisions decisions, a unified and objective framework to
simulate dispersal is missing.

% Step Selection Functions
Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful to
investigate \textit{transient} movement behavior during dispersal and landscape
connectivity.

% Current Study
Here, we showcase the use of integrated step selection functions as simple,
individual-based, and spatially explicit movement model to simulate dispersal of
the endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions, thereby parametrizing a fully
mechanistic movement model describing how dispersing wild dogs move through the
available landscape. Based on this model, we simulate 80'000 dispersers moving
across the extent of the KAZA-TFCA and generate a set of three connectivity
maps, each focused on a different aspect of connectivity. Finally, we discuss
the benefits and pitfalls of such a simulation-based approach and highlight
potential improvements to be made in the future.

@@

\section{Introduction}

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the social structure of wild animal populations that are
distributed in space \citep{Hanski.1998, Clobert.2012} and may strongly affect
population dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to maintain genetic diversity
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013}, to
rescue small, non-viable populations \citep{Brown.1977}, and to promote the
colonization or recolonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
landscape connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become a
fundamental task in conservation science \citep{Nathan.2008, Doerr.2011,
Rudnick.2012}. The ability to pinpoint relevant dispersal hotspots requires
information on movement behavior during dispersal and knowledge about factors
that limit dispersal and therefore connectivity \citep{Baguette.2013,
Vasudev.2015}.

% Advancements in GPS Technology & Movement Ecology
\subsection{Advancements in GPS Technology \& Movement Ecology (90\%)}
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the use of GPS data to study animal dispersal and
connectivity has accelerated \citep{Elliot.2014, Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded a ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the availability of large amounts of
empirical data and an increased computational power have led to the development
of numerous techniques to study dispersal and highlight critical corridors
between subpopulations \citep{Boyce.2002, Fortin.2005, Cushman.2010,
Zeller.2012, Diniz.2020}.

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory (CT;
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape to infer likely movement corridors. Despite their
intuitive nature and ease of use, both methods make rigorous assumptions about
animal movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be fulfilled by
migrating animals that typically move between a discrete set of habitats through
familiar landscapes. Dispersers, on the other hand, usually move over long
distances into unknown territory and are therefore less likely to be aware of
associated movement costs \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Another
issue of LCPs analysis concerns the fact that least-costly routes, by their very
nature, are only one pixel wide \citep{Pinto.2009}. This neglects the fact that
alternative routes with similar costs may exist and implies that the width of
inferred movement routes depends on the resolution of chosen covariate layers
and may not be biologically meaningful \citep{Diniz.2020}. Although some of
these deficiencies can be addressed using less stringent versions of the least
cost algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness remains.

% Issues with Circuit Theory
\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are rarely ever met. For
example, because CT only allows movements from a source cell to its four or
eight adjacent cells, it implicitly posits that animals exhibit a perceptual
range of a single pixel. Given that covariate layers are usually resolved with a
pixel size between 30 m x 30 m and 1 km x 1 km, this hardly ever renders the
true capability of animals to perceive the environment. Moreover, CT is built
around the assumption of a complete random walk \citep{Diniz.2020}, entailing
that directional biases cannot be rendered. Nevertheless, directionality is a
common characteristic in animal movement \citep{Bovet.1991, Schultz.2001},
especially in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}.

% Issues with Both Methods
\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal \citep{Diniz.2020}. Statements about the expected
duration required to traverse a certain corridor are therefore impossible.
Likewise, because movement is not modeled explicitly, interactions between
movement and habitat preferences of the focal species cannot be rendered.
Connectivity therefore merely arises in result to the landscape structure, which
is usually referred to as structural connectivity. While structural connectivity
yields insights in the \textit{potential} of the landscape to be traversed, it
does not enable to quantify the \textit{actual} gene flow through the area.
Consequently, a functional view on connectivity, which also renders the
behavioral response of the animal with respect to prevailing habitat conditions,
is often more desirable \citep{Tischendorf.2000, Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal trajectories are simulated spatially explicitly, based on movement
rules that determine how individuals move over and interact with the prevailing
landscape \citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004,
Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020, Vasudev.2021}. Using the simulated trajectories,
one can calculate a set of connectivity metrics, such as inter-patch
connectivity and traversal frequency, to reveal major dispersal corridors
\citep{Kanagaraj.2013, BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}.
However, while IBMMS can be employed to overcome any of the shortcomings
intrinsic to LCPs and CT, they are subject to a vast amount of subjective,
non-trivial modeling decisions. Moreover, they can be challenging to fit and
require vast amounts of movement data, ideally collected during dispersal
\citep{Diniz.2020}. Consequently, alternative methods that require fewer
modeling decisions and are straight forward to apply are desirable.

@@

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a basic habitat selection model and predicted a
permeability surface that we used to compute least-cost paths and corridors. We
now expand on this knowledge and use ISSFs to develop a more mechanistic
movement model of dispersing wild dogs (\Cref{GraphicalAbstract}). We employ the
model to simulate dispersers moving across the KAZA-TFCA and generate three
distinct connectivity maps, each shedding light onto a different aspect of
connectivity. With this work, we exemplify how ISSFs can be utilized for
dispersal simulations and we discuss several benefits of this approach over
traditional connectivity modeling techniques such as least-cost analysis and
circuit theory. Most importantly, simulations based on ISSFs provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addition, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis.

@@

\subsection{GPS Relocation Data (90\%)}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana \citep{Cozzi.2020, Hofmann.2021}. We selected potential
dispersers based on age, pack size, number of same‐sex siblings within the pack,
and presence of unrelated opposite-sex individuals in the pack
\citep{McNutt.1996, Behr.2020}. We immobilized selected individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020} that was
injected by dart, fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Immobilized individuals were fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}) that guaranteed automated drop-off through a decomposable piece of
cotton. Handling and collaring of all individuals was supervised by a
Botswana-registered wildlife veterinarian and all individuals quickly rejoined
their pack after immobilization.

16 collared individuals eventually dispersed, each in a separate same-sex
dispersal coalition (7 female and 9 male coalitions). During dispersal, collars
were programmed to record a GPS fix every 4 hours, all of which were regularly
transmitted over the Iridium satellite system, thereby allowing to remotely
track individuals, even if they left the main study area and crossed
international borders. Because behavior during dispersal is more pertinent for
assessing landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded
all data that was collected during residency and only retained GPS data recorded
during dispersal. In some instances, exact dispersal dates were known from field
observations, whereas in other cases we determined dispersal phases using the
net-squared displacement metric. Net squared displacement measures the squared
Euclidean distance of a GPS relocation to a reference point \citep{Borger.2012},
which in our case was set to the center of each individual's natal home range.
Thus, dispersal was deemed to have started when an individual left its natal
home range and ended once individuals became sedentary again. As previous
research found no differences in behaviors of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by an individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not successfully collected
on the 4-hourly schedule (\( \pm \) 15 minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting water layers thus temporally aligned with our
dispersal events. We furthermore computed a proxy for human influence, rendering
anthropogenic pressures stemming from human-density, agricultural sites, and
roads. All spatial layers were coarsened or interpolated to a target resolution
of 250 m by 250 m. Further details on the sources and preparation of each
habitat covariate are given in \cite{Hofmann.2021}.

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed step plus its 24 random steps formed a stratum
and received a unique identifier. As suggested by \cite{Avgar.2016}, we
generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale = 6'308, shape = 0.37). Along each step, we
extracted and averaged spatial covariates using the {\tt velox} package
\citep{Hunziker.2021}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each observed and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| > 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

@@

\subsection{Dispersal Simulation (80\%)}
We used the most parsimonious movement model to simulate 80'000 virtual
dispersers moving across the KAZA-TFCA. The simulation resembled an inverted
ISSF and was set up as follows. (1) We defined a random source point and assumed
a random initial orientation of the animal. (2) Departing from the source point,
we generated 25 random steps by sampling turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from our fitted gamma
distribution. Similar to the input data, each random step represented the
straight line movement within 4 hours. To prevent unreasonably large steps, we
capped sampled step lengths to a maximum of 35 km, which corresponded to the
farthest distance ever traveled within 4 hours by one of our dispersers. (3)
Along each random step, we extracted and averaged habitat covariates and we
calculated movement covariates. To ensure compatible scales, we standardized
extracted values using the same parameters applied to our input data. (4) We
applied the parametrized movement model to predict the selection score \(w(x)\)
for each step and we translated predicted scores into probabilities using
\Cref{EQ2}. (5) We sampled one of the random steps based on predicted
probabilities and determined the animal's new position. We repeated steps (2) to
(5) until 2'000 steps were realized, implying a total 160 Mio. simulated steps.

@@

\subsection{Source Points (90\%)}
We released 80'000 virtual dispersers from 80'000 unique source points
distributed across the study area. 50'000 virtual dispersers were released from
randomly selected source points within contiguous protected areas larger \(>\)
700 km\textsuperscript{2} (\Cref{SourcePoints}a), which conforms to average home
range requirements of resident wild dogs \citep{Pomilia.2015} and allowed us to
remove patches too small to host viable populations. By distributing source
points randomly, the number of source points per km\textsuperscript{2} was
approximately equal within protected areas. To render potential immigrants into
the study system, we released another 30'000 dispersers at random locations
inside the 100 km wide buffer zone surrounding the main study area
(\Cref{SourcePoints}b).

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourceAreas.png}
    \caption{(a) Different source areas from which we released virtual
    dispersers. We only considered contiguous protected areas (national parks,
    game reserves, and forest reserves) that were larger than 700
    km\textsuperscript{2} (green). This area corresponds to the average home
    range requirement for viable wild dog populations \citep{Pomilia.2015}. To
    render potential immigrants into the study system, we also initiated
    dispersers within a buffer zone (blue) surrounding the main study area. (b)
    Source points from which dipsersers were released. 50'000 dispersers were
    released from the main study area (green dots) and another 30'000 dispersers
    within the virtual buffer (blue dots).}
    \label{SourcePoints}
  \end{center}
\end{figure}

@@

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers \citep{Hauenstein.2019, Peer.2008}. For this,
we rasterized all simulated trajectories and tallied them into a single map. If
the same trajectory crossed a raster-cell twice, we only counted it once,
thereby mitigating potential biases caused by individuals that were trapped and
moved in circles. To achieve high performance rasterization, we used the
R-package {\tt terra} \citep{Hijmans.2021}.

@@

\subsection{Inter-Patch Connectivity (80\%)}
We assessed inter-patch connectivity between national parks located in our study
area to examine functional links between distinct patches in the KAZA-TFCA. The
decision to focus on national parks was purely out of simplicity and does not
imply that connections between other regions are impossible. In fact, the same
logic could easily be expanded to include other protected areas. To quantify
inter-patch connectivity, we computed the relative frequency at which dispersers
originating from one national park successfully moved into another national
park. Successful movement was said to be achieved if the individuals' trajectory
intersected with the corresponding national park at least once. We also recorded
the number of steps required until the first intersection with the polygon of
the respective national park. This allowed us to determine \textit{if} and
\textit{how often} dispersers moved between certain national parks, as well as
\textit{how long} dispersers had to move to realize those connections.

@@

Assuming that all other covariates are held constant at their means, the habitat
kernel reveals that dispersing wild dogs avoid water but prefer its proximity.
Similarly, dispersers avoid areas that are covered by woodlands, yet prefer
regions covered by shrublands or grasslands. Finally, dispersers avoid movement
through landscapes that are dominated by humans. Effect sizes are strong and,
except for effect of \textsf{distance to water}, statistically clear on the 5\%
significance level.

@@

\subsection{Dispersal Simulation (80\%)}
On a machine with an octacore AMD Ryzen 7 2700X processor (8 x 3.6 GHz) and 64
GB of RAM, a batch of 1'000 simulated dispersers moving over 2'000 steps
required 90 minutes to compute (\(\mu = 88.90\), \(\sigma = 1.87\)).
Consequently, the simulation of all 80'000 dispersers (160 Mio. steps)
terminated after 120 hours (i.e. five days). Comparable simulations will be
substantially faster for smaller study areas and lower resolution covariates, as
the covariate extraction from large and high-resolution rasters was
computationally the most demanding task. Out of the 50'000 dispersers initiated
inside the main source area \Cref{SourcePoints}(a), only 4.5\% eventually hit a
map boundary, suggesting that we successfully prevented biases due to boundary
effects. In contrast, 78\% of the 30'000 dispersers originating from the buffer
zone eventually hit a map boundary, yet this was to be expected since many of
those dispersers originated from source points located close to the map
boundary.
